### 2019-12-09 RFCs/HDF5/VOL/connector_author_guide/connector_interface.tex
The VOL connector author guide should be largely done for the time being.
Former-commit-id: 45242fb08a77b641dbe164270db51dcb6dc0d19c

The callback interface defined for the VOL has to be general enough to
handle all the HDF5 API operations that would access the
file. Furthermore, it has to capture future additions to the HDF5
library with little to no changes to the callback interface. Changing
the interface often whenever new features are added would be
discouraging to connector developers since that would mean reworking
their VOL connector structure. To remedy this issue, every callback will
contain two parameters:
\begin{itemize}
\item A data transfer property list (DXPL) which allows that API to
  put some properties on for the connectors to retrieve if they have to
  for particular operations, without having to add arguments to the
  VOL callback function.
\item A pointer to a request (\texttt{void **req}) to handle asynchronous
  operations if the HDF5 library adds support for them in future
  releases (beyond the 1.8 series). That pointer is set by the VOL
  connector to a request object it creates to manage progress on that
  asynchronous operation. If the \texttt{req} is \texttt{NULL}, that means
  that the API operation is blocking and so the connector would not
  execute the operation asynchronously. If the connector does not support
  asynchronous operations, it needs not to worry about this field and
  leaves it unset.
\end{itemize}

In order to keep the number of the VOL object classes and callbacks
concise and readable, it was decided not to have a one-to-one mapping
between API operation and callbacks. Furthermore, to keep the
callbacks themselves short and not cluttered with a lot of parameters,
some of the parameters are passed in as properties in property lists
included with the callback. The value of those properties can be
retrieved by calling the public routine (or its private version if
this is an internal connector): 
\begin{lstlisting}
herr_t H5Pget(hid_t plist_id, const char *property_name, void *value);
\end{lstlisting}
The property names and value types will be detailed when describing
each callback in their respective sections.

@@

\newpage

\begin{lstlisting}[caption={Structure to hold parameters for object locations, H5VLpublic.h}, captionpos=b, label={lst:ObjLocclass}]
/* 
 * Structure to hold parameters for object locations.
 * either: BY_SELF, BY_NAME, BY_IDX, BY_TOKEN
 */

@@

\newpage

Other API routines that would make a one-to-one mapping difficult are:
\begin{itemize} 
\item the \texttt{Get} operations that retrieve something from an object; for example a property list or a datatype of a dataset, etc.
\item API routines are specific to each class of callbacks (attributes, files, etc...). Having a callback for each one would clutter the VOL callback structure and reduce user-friendliness. 
\item API routines that make sense to only implement in the native HDF5 file format and would not be possible to implement in other connectors. Adding a callback for those routines would be confusing to connector developers. 
\end{itemize}

>>>>

The callback interface defined for the VOL has to be general enough to
handle all the HDF5 API operations that would access the
file. Furthermore, it has to capture future additions to the HDF5
library with little to no changes to the callback interface. Changing
the interface often whenever new features are added would be
discouraging to connector developers since that would mean reworking
their VOL connector structure. To remedy this issue, every callback will
contain two parameters:
\begin{itemize}
\item A data transfer property list (DXPL) which allows that API to
  put some properties on for the connectors to retrieve if they have to
  for particular operations, without having to add arguments to the
  VOL callback function.
\item A pointer to a request (\texttt{void **req}) to handle asynchronous
  operations if the HDF5 library adds support for them in future
  releases. That pointer is set by the VOL
  connector to a request object it creates to manage progress on that
  asynchronous operation. If the \texttt{req} is \texttt{NULL}, that means
  that the API operation is blocking and so the connector would not
  execute the operation asynchronously. If the connector does not support
  asynchronous operations, it needs not to worry about this field and
  leaves it unset.
\end{itemize}

In order to keep the number of the VOL object classes and callbacks
concise and readable, it was decided not to have a one-to-one mapping
between API operation and callbacks. The parameter names and types will be
detailed when describing each callback in their respective sections.

@@

\begin{lstlisting}[caption={Structure to hold parameters for object locations, H5VLconnector.h}, captionpos=b, label={lst:ObjLocclass}]
/* 
 * Structure to hold parameters for object locations.
 * either: BY_SELF, BY_NAME, BY_IDX, BY_TOKEN
 */

@@

### 2019-12-09 RFCs/HDF5/VOL/connector_author_guide/intro.tex
The VOL connector author guide should be largely done for the time being.
Former-commit-id: 45242fb08a77b641dbe164270db51dcb6dc0d19c

This guide is for people who are interested in developing their own VOL
connector for the HDF5 library. It is assumed that the reader has good knowledge
of the VOL architecture obtained by reading the VOL architectural design
document~\cite{vol:rfc}.

>>>>

This guide is for people who are interested in developing their own VOL
connector for the HDF5 library. It is assumed that the reader has good knowledge
of the VOL architecture obtained by reading the VOL architectural design
document.

This is a preliminary version of this document. The final version will be
released with the HDF5 1.12.0 release in late 2019.

### 2019-12-09 RFCs/HDF5/VOL/connector_author_guide/new_api_calls.tex
The VOL connector author guide should be largely done for the time being.
Former-commit-id: 45242fb08a77b641dbe164270db51dcb6dc0d19c

\section{New VOL API Routines}
\label{sec:api}
New API routines have been added to the HDF5 library to manage VOL connectors. This section details each new API call and explains their intended usage.

Additionally, a set of API calls that map directly to the VOL callbacks themselves have been added to aid in the development of passthrough connectors which can be stacked and/or split. For more information on usage refer to Section~\ref{sec:stack}. The routines are listed in Appendix~\ref{sec:B}.

@@

@@

\subsubsection{H5VLunregister\_connector}
\begin{mdframed}[style=bgbox]
\textbf{Signature:}
\begin{lstlisting}
herr_t H5VLunregister(hid_t connector_id);
\end{lstlisting}
\textbf{Arguments:}\\
\begin{tabular}{l p{13.5cm}}
  {\tt connector\_id} & (IN): A valid identifier of the connector to unregister.\\
\end{tabular}
\end{mdframed}
Unregisters a connector from the library and return a positive value on success otherwise return a negative value. That native VOL connector cannot be unregistered (this will return a negative herr\_t value).\bigskip

@@

\subsubsection{H5VLobject\_register}
\begin{mdframed}[style=bgbox]
\textbf{Signature:}
\begin{lstlisting}
  hid_t H5VLobject_register(void *obj, H5I_type_t obj_type, hid_t connector_id);
\end{lstlisting}

\textbf{Arguments:}\\
\begin{tabular}{l p{13.5cm}}
 {\tt obj} & (IN): pointer to the object to register.\\
 {\tt obj\_type} & (IN): The type of the object ({\tt H5I\_FILE, H5I\_DATASET, H5I\_GROUP, H5I\_ATTRIBUTE, H5I\_DATATYPE}).\\
 {\tt connector\_id} & (IN): identifier of the vol connector the object is associated with.\\
\end{tabular}
\end{mdframed}
Creates an HDF5 identifier for an object given its type and an identifier of the connector that was used in creating it. This function is typically used by external connectors wanting to wrap an {\tt hid\_t} around a connector specific object to return to a user callback for functions like {\tt H5Literate, H5Lvisit, H5Ovisit, H5Aiterate} etc... A Negative value is returned on failure.\bigskip

\subsubsection{H5VLget\_object}
\begin{mdframed}[style=bgbox]
\textbf{Signature:}
\begin{lstlisting}
  herr_t H5VLget_object(hid_t obj_id, void **obj); 
\end{lstlisting}

\textbf{Arguments:}\\
\begin{tabular}{l p{13.5cm}}
 {\tt obj\_id} & (IN): identifier of the object to dereference.\\
 {\tt obj} & (OUT): buffer to store the pointer of the VOL object associated with the object identifier.\\
\end{tabular}
\end{mdframed}
Retrieves a pointer to the VOL object from an HDF5 file or object identifier.\bigskip

>>>>

\section{New VOL API Routines}
\label{sec:api}
New API routines have been added to the HDF5 library to manage VOL connectors. This section details each new API call and explains their intended usage. Additionally, a set of API calls that map directly to the VOL callbacks themselves have been added to aid in the development of passthrough connectors which can be stacked and/or split. A list of these API calls is given in an appendix.

@@

The API calls in this header are for VOL management and general use (i.e., not
limited to VOL connector authors).

@@

\subsubsection{H5VLunregister\_connector}
\begin{mdframed}[style=bgbox]
\textbf{Signature:}
\begin{lstlisting}
herr_t H5VLunregister(hid_t connector_id);
\end{lstlisting}
\textbf{Arguments:}\\
\begin{tabular}{l p{13.5cm}}
  {\tt connector\_id} & (IN): A valid identifier of the connector to unregister.\\
\end{tabular}
\end{mdframed}
Unregisters a connector from the library and return a positive value on success otherwise return a negative value. The native VOL connector cannot be unregistered (this will return a negative herr\_t value).\bigskip

@@

This functionality is intended for VOL connector authors and includes helper
functions that are useful for writing connectors. Callback equivalent
functions can be found in {\tt H5VLconnector\_passthru.h}. A list of these
functions is included as an appendix to this document.

\subsubsection{H5VLobject}
\begin{mdframed}[style=bgbox]
\textbf{Signature:}
\begin{lstlisting}
  void *H5VLget_object(hid_t obj_id); 
\end{lstlisting}

\textbf{Arguments:}\\
\begin{tabular}{l p{13.5cm}}
 {\tt obj\_id} & (IN): identifier of the object to dereference.\\
\end{tabular}
\end{mdframed}
Retrieves a pointer to the VOL object from an HDF5 file or object identifier.\bigskip

\subsubsection{H5VLget\_file\_type}
\begin{mdframed}[style=bgbox]
\textbf{Signature:}
\begin{lstlisting}
  hid_t H5VLget_file_type(void *file_obj, hid_t connector_id, hid_t dtype_id); 
\end{lstlisting}

\textbf{Arguments:}\\
\begin{tabular}{l p{13.5cm}}
 {\tt file\_obj} & (IN): pointer to a file or file object's connector-specific data.\\
 {\tt connector\_id} & (IN): A valid identifier of the connector to use.\\
 {\tt dtype\_id} & (IN): A valid identifier for the type.\\
\end{tabular}
\end{mdframed}
Returns a copy of the {\tt dtype\_id} parameter but with the location set to
be in the file. Returns a negative value ({\tt H5I\_INVALID\_HID}) on errors.\bigskip

\subsubsection{H5VLpeek\_connector}
\begin{mdframed}[style=bgbox]
\textbf{Signature:}
\begin{lstlisting}
  hid_t H5VLpeek_connector_id(const char *name); 
\end{lstlisting}

\textbf{Arguments:}\\
\begin{tabular}{l p{13.5cm}}
 {\tt name} & (IN): name of the connector to query.\\
\end{tabular}
\end{mdframed}
Retrieves the ID for a registered VOL connector. This is done without duplicating
the ID and transferring ownership to the caller (as it normally the case in the
HDF5 library). The ID returned from this operation should not be closed.
This is intended for use by VOL connectors to find their own ID.

Returns a negative value ({\tt H5I\_INVALID\_HID}) on errors.\bigskip

### 2019-12-09 RFCs/HDF5/VOL/connector_author_guide/writing_new.tex
The VOL connector author guide should be largely done for the time being.
Former-commit-id: 45242fb08a77b641dbe164270db51dcb6dc0d19c

\section{Creating and Using an External Plugin}
External plugins are developed outside of the HDF5 library and do not
use any internal HDF5 private functions. They do not require to be
shipped with the HDF5 library, but can just link to it from userspace
just like an HDF5 application. If an external VOL plugin is installed on the system as a shared library or a DLL, the library can search for it and load it dynamically to be used by an application.

\subsection{Creating an External Plugin}
Developing an external plugin is similar to developing an internal plugin. Refer to Section~\ref{sec:vol} for details on plugin creation. The important thing to keep in mind is that external plugins cannot use internal HDF5 features. The ``value'' field for an external plugin should be a positive integer greater than 128. The name should be unique across all plugins registered with the library.

@@

Implementing iteration and visit callbacks in external plugins requires the plugin invoking the user defined callback function with an HDF5 identifier ({\tt hid\_t}) on the location object. The current HDF5 H5I API does not allow creating and dereferencing identifiers that point to HDF5 object types. Therefore, two new API routines have been added (see Section~\ref{sec:api} for more details) to get around that. {\tt H5VLobject\_register()} is used to create an HDF5 identifier for a VOL object handle (file, group, dataset, named datatype, or attribute). {\tt H5VLget\_object()} is used to derefernce the HDF5 identifier back to the VOL object handle.

\subsection{Using an External Plugin}
Unlike internal plugins, external plugins cannot create an API routine for applications to use to set the VOL plugin in the file access property list. External plugin developers can, however, provide a wrapper routine that registers the plugin with {\tt H5VLregister()} instead of having the application call that directly with the pointer to the plugin class structure. Both ways are possible if the external plugin used is in the application's userspace. External plugins provided by third party vendors that are not in the application's userspace but are installed on the system with a shared library or DLL can be registered by the application using {\tt H5VLregister\_name()} provided that the application knows the plugin's name in advance.

The registration operation will return a global identifier for the registered plugin. Applications can query the library for that identifier using the plugin name with {\tt H5VLget\_plugin\_id()}. This identifier is used to initialize the plugin if it requires an initialization phase. A third party library could use that ID to create a plugin that stacks or mirrors on top of an internal plugin using that identifier (refer to Section~\ref{sec:stack} for more details on stacking and mirroring).

Some plugins could require an initialization phase before using the plugin with some input parameters from users. This process is done by creating a VOL initialization property list and setting a public property that should be defined by the internal plugin to the hostname of the remote machine. The user then calls the initialize routine ({\tt H5VLintialize()}) with the plugin\_id and the property list to initialize the plugin. An external plugin could abstract those operations into a wrapper routine that wraps the property list creation, property settings, and plugin initialization into one call that the user makes to simplify this process.

The application then sets the plugin access property in the file access property list to the external plugin using the plugin identifier with this API routine:
\begin{lstlisting}
  herr_t H5Pset_vol(hid_t fapl_id, hid_t plugin_id, const void *new_vol_info);
\end{lstlisting}
where {\tt new\_vol\_info} is the plugin information needed from the application. Typically it is a plugin defined structure with some fields to setup the plugin access. The external plugin could also provide a wrapper function around this to make it easier for the application to use.

Now that the plugin is initialized, the application can create or open an HDF5 container using the external plugin. After closing all objects and containers on that plugin, the application should terminate access to the selected plugin if it requires a termination phase by calling {\tt H5VLterminate()} with the required termination properties (similar to the initialization phase).

Finally, the application is required to un-register the plugin from the library when access to the container(s) is terminated using {\tt H5VLunregister}.

% /* Function prototypes */
% H5_DLL htri_t H5VLis_registered(hid_t id);
% H5_DLL ssize_t H5VLget_plugin_name(hid_t id, char *name/*out*/, size_t size);

% H5_DLL void *H5Pget_vol_info(hid_t plist_id);
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End:

>>>>

\section{Creating a New Connector}

\subsection{Overview}
Creating a new VOL connector can be a complicated process. You will need to map
your storage system to the HDF5 data model through the lens of the VOL and this
may involve some impedence mismatch that you will have to work around. The good
news is that the HDF5 library has been re-engineered to handle arbitrary,
connector-specific data structures via the VOL callbacks, so no knowledge of
the library internals is necessary to write a VOL connector.

Writing a VOL connector requires these things:

\begin{enumerate}
    \item Decide on library vs plugin vs internal.
    \item Set up your build/test files (CMake, Autotools, etc.).
    \item Fill in some boilerplate information in your {\tt H5VL\_class\_t} struct.
    \item Decide how you will perform any necessary initialization needed
        by your storage system.
    \item Map Storage to HDF5 File Objects
    \item Create implementations for the callbacks you need to support.
    \item Test the connector.
\end{enumerate}

Each of the steps listed above is decribed in more detail in this section of the document.

The "model then implement" steps can be performed iteratively. You might begin
by only supporting files, datasets, and groups and only allowing basic operations
on them. In some cases, this may be all that is needed. As your needs grow, you
can repeat those steps and increase the connector's HDF5 API coverage at a pace
that makes sense for your users.

Also, note that this document only covers writing VOL connectors using the
C programming language. It is often possible to write connectors in other programming languages
(e.g.; Python) via the language's C interop facilities, but that topic is out of
scope for this document.

\subsection{Sources Of Information}

Use of the VOL, including topics such as registration and loading VOL plugins, is
described in the VOL User Guide.

Public header files you will need to be familiar with include:

\begin{tabularx}{\linewidth}{ l X }
  \texttt{H5VLpublic.h} & Public VOL header.\\
  \texttt{H5VLconnector.h} & Main connector author header. Contains definitions for the main VOL struct and callbacks, enum values, etc.\\
  \texttt{H5VLconnector\_passthru.h} & Helper routines for passthrough connector authors.\\
    \texttt{H5VLnative.h} & Native VOL connector header. May be useful if your connector will attempt to implement native HDF5 API calls that are handled via the \textit{optional} callbacks.\\
  \texttt{H5PLextern.h} & Needed if your connector will be built as a plugin.\\
\end{tabularx}

Many VOL connectors can be found on The HDF Group's Bitbucket server, located at
\url{https://bitbucket.hdfgroup.org/projects/HDF5VOL}. Not all of these VOL
connectors are supported by THe HDF Group and the level of completeness varies,
but the connectors found there can serve as examples of working implementations.

\subsection{Library vs Plugin vs Internal}

When building a VOL connector, you have several options:

\subsubsection*{Library}

The connector can be built as a normal shared or static library. Software that
uses your connector will have to link to it just like any other library. This
can be convenient since you don't have to deal with plugin paths and searching
for the connector at runtime, but it also means that software which uses
your connector will have to be built and linked against it.

\subsubsection*{Plugin}

You can also build your connector as a dynamically loaded plugin. The mechanism
for this is the same mechanism used to dynamically load HDF5 filter plugins. This can
allow use of your connector via the VOL environment variable, without modifying
the application, but requires your plugin to be discoverable at runtime. See
the VOL User Guide for more information about using HDF5 plugins.

To build your connector as a plugin, you will have to include {\tt H5PLextern.h}
(a public header distributed with the library) and implement the
{\tt H5PLget\_plugin\_type()} and {\tt H5PLget\_plugin\_info()}
calls, both of which are trivial to code up. It also often requires your
connector to be built with certain compile/link options. The VOL connector
template does all of these things.

To show how easy this is to accomplish, here is the complete implementation
of those functions in the template VOL connector:

\begin{lstlisting}
H5PL_type_t H5PLget_plugin_type(void) {return H5PL_TYPE_VOL;}
const void *H5PLget_plugin_info(void) {return &template_class_g;}
\end{lstlisting}

The HDF5 library's plugin loading code will call {\tt H5PLget\_plugin\_type()}
to determine the type of plugin (e.g.; filter, VOL) and {\tt H5PLget\_plugin\_info()}
to get the class struct, which allows the library to query the plugin for its
name and value to see if it has found a requested plugin. When a match is found,
the library will use the class struct to register the connector and map its callbacks.

@@

\subsubsection*{Internal}

Your VOL connector can also be constructed as a part of the HDF5 library. This
works in the same way as the stdio and multi virtual file drivers (VFDs) and
does not require knowledge of HDF5 internals or use of non-public API calls. You
simply have to add your connector's files to the {\tt Makefile.am} and/or
{\tt CMakeLists.txt} files in the source distribution's {\tt src} directory.
This requires maintaining a private build of the library, though, and is not
recommended.

\subsection{Build Files / VOL Template}

We have created a template VOL connector that includes both Autotools and CMake
build files. The constructed VOL connector includes no real functionality, but 
can be registered and loaded as a plugin.

The VOL template  can be found here:

\quad \quad \url{https://bitbucket.hdfgroup.org/projects/HDF5VOL/repos/template}

The purpose of this template is to quickly get you to the point where you can
begin filling in the callback functions and writing tests. You can copy this
code to your own repository to serve as the basis for your new connector.

\subsection{{\tt H5VL\_class\_t} Boilerplate}

Several fields in the {\tt H5VL\_class\_t} struct will need to be filled in.

In HDF5 1.12.0, {\tt version} will be 1, indicating version 1 of the {\tt H5VL\_class\_t} struct.

Every connector needs a {\tt name} and {\tt value}. The library will use these when
loading and registering the connector (as described in the VOL User Guide), so
they should be unique in your ecosystem.

VOL connector values are integers, with a maximum value of 65535. Values from
0 to 255 are reserved for internal use by The HDF Group. The native VOL connector
has a value of 0.

As is the case with HDF5 filters, The HDF Group can register the value you've
chosen for your VOL connector so that others don't use it. Please contact
\href{mailto:help@hdfgroup.org}{help@hdfgroup.org} for help with this. We currently
do not register connector names.

The {\tt cap\_flags} field is currently not fully utilized. In the future, this
will be most likely be used to indicate which callbacks and capabilities are
supported by the VOL connector, but for now this struct field contains few flags, 
a list of which can be found in {\tt H5VLconnector.h}.

\subsection{Initialization and Shutdown}

You'll need to decide how to perform any initialization and shutdown tasks that
are required by your connector. There are initialize and terminate callbacks
in the {\tt H5VL\_class\_t} struct to handle this. They are invoked when
the connector is registered and unregistered, respectively. If this is
unsuitable, you may have to create custom connector-specific API calls to handle
initialization and termination. It may also be useful to perform operations in
a custom API call used to set the VOL connector in the fapl.

\subsection{Map Storage to HDF5 File Objects}

The most difficult part of designing a new VOL connector is going to determining
how to support HDF5 file objects and operaions using your storage system.
There isn't much specific advice to give here, as
each connector will have unique needs, but a forthcoming "tutorial" connector
will set up a simple connector and demonstrate this process.

\subsection{Fill In VOL Callbacks}

For each file object you support in your connector (including the file itself),
you will need to create a data struct to hold whatever file object
metadata that are needed by your connector. For example, a data structure for
a VOL connector based on text files might have a file struct that contains a
file pointer for the text file, buffers used for cacheing data, etc. Pointers
to these data structures are where your connector's state is stored and are
returned to the HDF5 library from the create/open/etc. callbacks such as
\textit{dataset create}.

Once you have your data structures, you'll need to create your own
implementations of the callback functions and map them via your
{\tt H5VL\_class\_t} struct. 
\subsection{Testing Your Connector}

At the time of writing, some of the HDF5 library tests have been abstracted
out of the library with their native-file-format-only sections removed and
added to a VOL test suite available here:

\quad \quad \url{https://bitbucket.hdfgroup.org/projects/HDF5VOL/repos/vol-tests}

This is an evolving set of tests, so see the documentation in that repository
for instructions as to its use. You may want to clone and modify and/or
extend these tests for use with your own connector.

In the future, we plan to modify the HDF5 test suite that ships with the
library to use a future VOL capabilities flags scheme to selectively run
tests that a particular connector supports. As this is a large task, it may be
some time before that work is complete.

\subsection{Passthrough Connectors}

Coming Soon

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End:
